V4046\,Sgr high-resolution 1.3\,mm continuum ALMA observations: a disc with a thin ring

### 2020-10-02 mnras_template.tex
Add files via upload

To analyse the new ALMA data, we aim to reproduce the disc through a parametric model ran with the radiative transfer code \textsc{radmc3d} \citep{Dullemond_2012} so that the model would agree with the ALMA observation, a previous SPHERE-IRDIS observation and the Spectral Energy Distribution (SED). With the model representation of the dust density distribution and combined with a decomposition in polar coordinates of the ALMA image we can have better approximations of the radial structure of the disc. It is worth mentioning that previous models of V4046\,Sgr have been made \citep{Ru_z_Rodr_guez_2019, Rosenfeld_2013, 2019ApJ...882..160Q} but with the difference that our new reference observation of radio wavelengths has a sub-arcsecond resolution, allowing us to discover previously unseen underlying substructures of the large-dust grains disc.

In the following sections, first we describe the new 1.3\,mm continuum observation alongside with a SPHERE-IRDIS \textit{H}-band image (Section \ref{sec:Observations}), present the structure of the parametric model used to simulate the disc (Section \ref{sec:model}), comment on the results of the comparison of the model with the data and a discussion of the main results (Section \ref{sec:results}). Finally, we share our main conclusions in Section \ref{sec:Conclusions}.

@@

Disc structures can be explored by means of comparing mm continuum and scattered light images. These observations allow us to see the distribution of different populations of dust particles, while ALMA traces the millimetre-sized grains settled in the mid-plane of the disc, scattered light accounts for photons reflected off small micron-sized dust at the disc surface layer, high above the mid-plane. By comparing both we can determine whether asymmetries in scattered light correspond to surface ripples or deep structures intrinsic to the underlying density distribution.

New ALMA observations of V4046\,Sgr were obtained in 2017 as part of the Cycle 5 program 2017.1.01167.S.. These were taken in the context of a larger survey for 14 optically visible TTauri discs at $\sim$50\,mas resolution. The survey simultaneously mapped 1.3\,mm continuum and the J = 2$-$1 line of $^{12}$CO with the C43-8 array configuration in band 6 (211-275\,GHz). In this work we focus exclusively on the data obtained in the 1.3\,mm continuum for our source, taken with a beam size of $0\farcs04 \,\times \, 0\farcs05$.

For this investigation we made use of previous polarimetric image (at 1.65$\micron$) taken at the ESO Very Large Telescope with the SPHERE-IRDIS instrument. A complete description of these image and all about data reduction was reported in \citet{Avenhaus_2018}.

For our analysis of the ALMA image we used an image reconstruction strategy, using the non-parametric image synthesis of the \textsc{uvmem} package \citep{2006ApJ...639..951C, 2018A&C....22...16C} we can super-resolve the clean beam, obtaining an effective angular resolution $\sim$2-3 times higher than in the clean image with natural weights. The model obtains this higher resolution through fitting the data using Chi-squared minimisation. Here we adopted a pure $\chi ^2$ model image. The resulting \textsc{uvmem} image shown in Fig. \ref{fig:two} exhibits previously unseen substructure of the disc. This unique high-definition image reveals that the disc features a two-ring structure for the large grain dust with a broad gap between the rings. The wide and brighter outer ring has been observed in previous occasions, but the thin and weak inner ring has never seen before (\citet{Ru_z_Rodr_guez_2019} anticipated its existence). The outer ring has its peak intensity at $\sim$30\,au and reaches out to $\sim$60\,au. Additionally, we can observe that there is a surface brightness break at $\sim$35\,au, another feature that has never been considered.

@@

Assuming a Gaussian dust inner ring, we obtained the mean width and the location of the ring by finding for each azimuth the Gaussian intensity profile that best describes the ring. This gave us a FWHM of 4.17$\pm$0.94\,au and that it is located at 13.46$\pm$0.43\,au from the stars (See Fig. \ref{fig:polarring}). As we are using the \textsc{uvmem} model image and the effective angular resolution is $\sim$1/3 that of the natural-weighted beam ($0\farcs04 \,\times \, 0\farcs05$), these measurements only gives as an upper limit of the real width.

@@

\citet{dOrazi} also measured the binary phase in the scattered light observation around 106\,$\degr$, setting $\phi=$0 from the North and increasing clockwise. We can then calculate the binary phase at the time of the ALMA observation at $\phi=$13\,$\degr$. There is no clear signal of any shadow similar to the ones present in the DPI image, this could be explained due to the slower cooling of larger grains (cite Casassus?***).

@@

Recreating the complex radial and vertical structure of the V4046\,Sgr disc is not an easy task. We choose to build the model by thinking the components of the disc separately and then unite them as a whole. Given the observations, we consider that a correct approach for this disc is using two different dust populations: larger grains that are vertically settled and dominates the disc mass and a less-abundant population of smaller grains that are distributed to larger heights from the mid-plane. As the gas and the small-dust population are coupled, it is important to consider the gas in the modelling process, but we are not predicting line emissions.

Assuming a three dimensional model in a spherical reference frame with coordinates (r, $\theta$, $\phi$), gas density ($\rho_{\mathrm{gas}}$) distribution follows
\begin{equation}
  \rho_{\mathrm{gas}}(r,z) =\frac{\Sigma(r) \,\delta(r)}{\sqrt{2\pi} \,r H(r)}  \text{exp}\left[-\frac{1}{2} \left(\frac{z}{r H(r)}\right)^2\right].
\end{equation}
Where $\delta(r)$ is a parameter that indicates the density drops in the gaps and cavities, $H(r)$ is the scale height profile and $\Sigma(r)$ been the surface density profile. $\Sigma(r)$ is defined by
\begin{equation}
  \Sigma(r) = \Sigma_\mathrm{c} \left(\frac{r}{R_\mathrm{c}}\right)^{-\gamma}  \, \text{exp}\left[-\left(\frac{r}{R_\mathrm{c}}\right)^{2-\gamma}\right],
\end{equation}
where $R_c$ is a characteristic radius and $\gamma$ is the surface density gradient. We used a fixed $\gamma$ = 1, a typical value for discs \citep{Andrews_2009,Andrews_2010}.

@@

Since the large-dust grains are less coupled to the gas, their distribution has some important differences. We leave a large inner cavity and added another scaling factor (or filter) to generate the larger gap between the rings. For the inner ring we use the same profile as the gas but with different values for $R_c$ and $\Sigma_c$. The greatest difference is in the outer ring, where we used a sum of an exponential correction with a Gaussian function and, between 32 and 34.5\,au, in an effort to recreate the break seen in the profile, we chose a different value for $\gamma$ = -8. Then we have

@@

Also for both dust populations, the inner part of the outer ring follows a different behaviour. To model that parts we multiplied by an additional factor $\epsilon(r)$
\begin{equation}
    \epsilon(r) = \delta +  (1 - \delta) \left(\frac{ r - R_\mathrm{in}}{R_\mathrm{peak} - R_\mathrm{in}}\right)^3,
\end{equation}
where $\delta=10^{-4}$ and $R_\mathrm{in}$ and $R_\mathrm{peak}$ would depend on the dust population, marking the beginning and the maximum density peak of the outer ring. This parameter allow us to model a smoother inner wall of the outer ring.

@@

For the vertical structure, \citet{dOrazi} found flaring angles of $\varphi$ = 6.2$\pm$0.6\,$\degr$ for the inner ring and $\varphi$ = 8.5$\pm$1.0\,$\degr$ for the outer one. For that matter, we used two different flaring index $\psi$ for our model. The separation between the two values was set at $r$ = 19\,au with values of 0.2 and 0.5 for the inner part and the outer part respectively. Also, the scale height is calculated using $H_o$ = 0.045 and $r_0$ = 19.5\,au.

We set the values of the inclination an the position angle as the same as those obtained from the ALMA observation in Section \ref{sec:Observations}, so the model has an inclination of i = 33.9\,$\degr$ and a P.A. = 74.6\,$\degr$. Finally, we set the distance at d = 72.4\,pc \citep{Gaia}.

@@

The simulated 1.3\,mm continuum image displays some clear similarities with the ALMA observation. The model reproduced the two rings that are present in the observation: the faint inner ring and the outer brighter one. %The outer ring has an intensity peak at $\sim$30\,au and a break present at $\sim$35\,au. 
As the radial profiles obtained from the simulated images of the model closely resemble those deduced from the observations (Fig. \ref{fig:radprofiles}), we can assume that the model provides a reliable approximation of the disc structure, including, particularly for our interest, the dimensions of the previously unseen 1.3\,mm thin inner ring. Accordingly, taking the parametric model values for the large-dust inner ring we can have that it is located at 13.6\,au, has an estimated width of 2.6\,au, an estimated scale height ranging from 0.25 to 0.32\,au and a dust mass of about 3\,M$_{\earth}$.

>>>>

To analyse the new ALMA data, we aim to reproduce the disc through a parametric model ran with the radiative transfer code \textsc{radmc3d} \citep{Dullemond_2012} so that the model would agree with the ALMA observation, a previous SPHERE-IRDIS observation and the Spectral Energy Distribution (SED). Combining the model representation of the dust density distribution with a decomposition in polar coordinates of the ALMA image we can have better approximations of the radial structure of the disc. It is worth mentioning that previous models of V4046\,Sgr have been made \citep{Ru_z_Rodr_guez_2019, Rosenfeld_2013, 2019ApJ...882..160Q} but with the difference that our new reference observation of radio wavelengths has a sub-arcsecond resolution, allowing us to discover previously unseen underlying substructures of the large-dust grains disc.

The structure of this letter is as follows. The new 1.3\,mm continuum observation is described alongside with a SPHERE-IRDIS \textit{H}-band image in Section \ref{sec:Observations}. The structure of the parametric model used to simulate the disc is presented in Section \ref{sec:model}. The results of the model and the discussion are commented Section \ref{sec:results} and the main conclusions are described in Section \ref{sec:Conclusions}.

@@

Disc structures can be explored by means of comparing mm continuum and scattered light images. These observations allow us to see the distribution of different populations of dust particles, while ALMA traces the millimetre-sized grains settled in the mid-plane of the disc, scattered light accounts for photons reflected off small micron-sized dust at the disc surface layer, high above the mid-plane. By comparing both is easier to determine whether asymmetries in scattered light correspond to surface ripples or deep structures intrinsic to the underlying density distribution.

New ALMA observations of V4046\,Sgr were obtained in 2017 as part of the Cycle 5 program 2017.1.01167.S.. These were taken in the context of a larger survey for 14 optically visible TTauri discs at $\sim$50\,mas resolution. The survey simultaneously mapped 1.3\,mm continuum and the J = 2$-$1 line of $^{12}$CO with the C43-8 array configuration in band 6 (211-275\,GHz). This work focuses exclusively on the data obtained in the 1.3\,mm continuum for our source, taken with a beam size of $0\farcs04 \,\times \, 0\farcs05$.

In this investigation, a previous polarimetric image at 1.65\,$\micron$ is used. The image was taken at the ESO Very Large Telescope with the SPHERE-IRDIS instrument, a complete description of the image and all about data reduction was reported in \citet{Avenhaus_2018}.

For our analysis of the ALMA image we used an image reconstruction strategy, using the non-parametric image synthesis of the \textsc{uvmem} package \citep{2006ApJ...639..951C, 2018A&C....22...16C} it is possible to super-resolve the clean beam, obtaining an effective angular resolution $\sim$2-3 times higher than in the clean image with natural weights. The model obtains this higher resolution through fitting the data using Chi-squared minimisation. Here we adopted a pure $\chi ^2$ model image. The resulting \textsc{uvmem} image shown in Fig. \ref{fig:two} exhibits previously unseen substructure of the disc. This unique high-definition image reveals that the disc features a two-ring structure for the large grain dust with a broad gap between the rings. The wide and brighter outer ring has been observed in previous occasions, but the thin and weak inner ring has never seen before (\citet{Ru_z_Rodr_guez_2019} anticipated its existence). The outer ring has its peak intensity at $\sim$30\,au and reaches out to $\sim$60\,au. Additionally, there is a surface brightness break at $\sim$35\,au, another feature that has never been considered.

@@

Assuming a Gaussian dust inner ring, we obtained the mean width and the location of the ring by finding for each azimuth the Gaussian intensity profile that best describes the ring. This gave us a FWHM of 4.17$\pm$0.94\,au and that it is located at 13.46$\pm$0.43\,au from the stars (See Fig. \ref{fig:polarring}). As the \textsc{uvmem} model image has an effective angular resolution of $\sim$1/3 that of the natural-weighted beam ($0\farcs04 \,\times \, 0\farcs05$), these measurements only gives an upper limit of the real width.

@@

\citet{dOrazi} also measured the binary phase in the scattered light observation around 106\,$\degr$, setting $\phi=$0 from the North and increasing clockwise. Using this measurement, the binary phase was calculated at the time of the ALMA observation at $\phi=$13\,$\degr$. There is no clear signal of any shadow similar to the ones present in the DPI image, this could be explained due to the slower cooling of larger grains (cite Casassus?***).

@@

Recreating the complex radial and vertical structure of the V4046\,Sgr disc is not an easy task. We choose to build the model by thinking the components of the disc separately and then unite them as a whole. Given the observations, we consider that a correct approach for this disc is using two different dust populations: larger grains that are vertically settled and dominates the disc mass and a less-abundant population of smaller grains that are distributed to larger heights from the mid-plane. As the gas and the small-dust population are coupled, it is important to consider the gas in the modelling process, but no line emission prediction is made.

Assuming a three dimensional model in a spherical reference frame with coordinates (r, $\theta$, $\phi$), gas density ($\rho_{\mathrm{gas}}$) distribution follows
\begin{equation}
  \rho_{\mathrm{gas}}(r,z) =\frac{\Sigma(r) \,\delta(r)}{\sqrt{2\pi} \,r H(r)}  \text{exp}\left[-\frac{1}{2} \left(\frac{z}{r H(r)}\right)^2\right].
\end{equation}
Where $\delta(r)$ is a parameter that indicates the density drops in the gaps and cavities, $H(r)$ is the scale height profile and $\Sigma(r)$ been the surface density profile. $\Sigma(r)$ is defined by
\begin{equation}
  \Sigma(r) = \Sigma_\mathrm{c} \left(\frac{r}{R_\mathrm{c}}\right)^{-\gamma}  \, \text{exp}\left[-\left(\frac{r}{R_\mathrm{c}}\right)^{2-\gamma}\right],
\end{equation}
where $R_c$ is a characteristic radius and $\gamma$ is the surface density gradient. A fixed $\gamma$ = 1 is used as it is a typical value for discs \citep{Andrews_2009,Andrews_2010}.

@@

Since the large-dust grains are less coupled to the gas, their distribution has some important differences. A large inner cavity and another scaling factor (or filter) to generate the larger gap between the rings are needed. For the inner ring, the same profile as the gas is used, but with different values for $R_c$ and $\Sigma_c$. The greatest difference is in the outer ring, where we used a sum of an exponential correction with a Gaussian function and, between 32 and 34.5\,au, in an effort to recreate the break seen in the profile, we chose a different value for $\gamma$ = -8. So, for the large dust grains, the surface density is

@@

Also for both dust populations, the inner part of the outer ring follows a different behaviour. The surface density profile is multiplied by an additional factor
\begin{equation}
    \epsilon(r) = \delta +  (1 - \delta) \left(\frac{ r - R_\mathrm{in}}{R_\mathrm{peak} - R_\mathrm{in}}\right)^3,
\end{equation}
where $\delta=10^{-4}$ and $R_\mathrm{in}$ and $R_\mathrm{peak}$ would depend on the dust population, marking the beginning and the maximum density peak of the outer ring. This parameter allow us to model a smoother inner wall of the outer ring.

@@

For the vertical structure, \citet{dOrazi} found flaring angles of $\varphi$ = 6.2$\pm$0.6\,$\degr$ for the inner ring and $\varphi$ = 8.5$\pm$1.0\,$\degr$ for the outer one. For that matter, the model uses two different flaring index $\psi$. The separation between the two values was set at $r$ = 19\,au with values of 0.2 and 0.5 for the inner part and the outer part respectively. Also, the scale height is calculated using $H_o$ = 0.045 and $r_0$ = 19.5\,au.

We set the values of the inclination an the position angle as the same as those obtained from the ALMA observation in Section \ref{sec:Observations}, so the model has an inclination of i = 33.9\,$\degr$ and a P.A. = 74.6\,$\degr$. Finally, the distance is set at d = 72.4\,pc \citep{Gaia}.

@@

The simulated 1.3\,mm continuum image displays some clear similarities with the ALMA observation. The model reproduced the two rings that are present in the observation: the faint inner ring and the outer brighter one. %The outer ring has an intensity peak at $\sim$30\,au and a break present at $\sim$35\,au. 
As the radial profiles obtained from the simulated images of the model closely resemble those deduced from the observations (Fig. \ref{fig:radprofiles}), we can assume that the model provides a reliable approximation of the disc structure, including, particularly for our interest, the dimensions of the previously unseen 1.3\,mm thin inner ring. Accordingly, taking the parametric model values for the large-dust inner ring we have that it is located at 13.6\,au, has an estimated width of 2.6\,au, an estimated scale height ranging from 0.25 to 0.32\,au and a dust mass of about 3\,M$_{\earth}$.
